Descriptive Complexity for Counting Complexity Classes

### 2018-12-27 LMCS/sendrev/framework.tex
carpeta con version send corregida

The syntax of QSO formulae is divided in two levels. 
The first level is composed by $\so$-formulae over $\R$ (called Boolean formulae) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called Boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, and $\Pi x$ and $\Pi X$ are called first- and second-order product, respectively.
This division between Boolean and quantitative level is essential for understanding the difference between the logic and the quantitative part. 
Furthermore, this will later allow us to parametrize both levels of the logic in order to capture different counting complexity classes.

@@

Suppose now that we want to count the number of cliques in~$\fG$. We can define this function with the following formula:
\[
\alpha_2 \ := \ \sa{X} \clique(X),
\] 
where $\clique(X) := \fa{x} \fa{y} ((X(x) \wedge X(y) \wedge x \neq y)  \rightarrow E(x,y))$.
In the Boolean sub-formula of $\alpha_2$ we check whether $X$ is a clique, and with the sum quantifier we add one for each clique in $\fG$. 
But in contrast to $\alpha_1$, 
in $\alpha_2$ we need a second-order quantifier in the quantitative level.
This is according to the
complexity of evaluating each formula:
$\alpha_1$ defines an $\fp$-function while $\alpha_2$ defines a $\shp$-complete function. \qed
\end{exa}
\begin{exa}\label{exa-perm}
For an example that includes multiplication, let $\bM = \{M(\cdot,\cdot),<\}$ be a vocabulary for storing 0-1 matrices; in particular, a structure $\fM$ over $\bM$ encodes a 0-1 matrix $A$ as follows: if $A[i,j] = 1$, then $M(i,j)$ is true, otherwise $M(i,j)$ is false.
Suppose now that we want to compute the permanent of an $n$-by-$n$ 0-1 matrix $A$, that is:
\begin{align*}
\op{perm}(A) &= \sum_{\sigma \in S_n} \prod_{i=1}^n A[i, \sigma(i)],  
\end{align*}
where $S_n$ is the set of all permutations over $\{1, \ldots, n\}$.
The permanent is a fundamental function on matrices that has found many applications;
% in different areas,
%~\cite{permanent-applications},
in fact, showing that this function is hard to compute was one of the main motivations behind the definition of the class $\shp$~\cite{Valiant79}.
%and it was one of the first function that was shown to be difficult for counting~\cite{Valiant79} (i.e. $\shp$-complete).

To define the permanent of a 0-1 matrix in $\qso$, assume that for a binary relation symbol $S$, $\op{permut}(S)$ is an $\fo$-formula that is true if, and only if, $S$ is a permutation, namely, a total bijective function (the definition of $\op{permut}(S)$ is straightforward).
%\martin{abreviar $\op{permutation}$ a $\op{permut}$ ahorra un par de lineas}
Then the following is a $\qso$-formula defining the permanent of a matrix:
\[
\alpha_3 := \sa{S} \op{permut}(S) \cdot \pa{x} (\ex{y} S(x,y) \wedge M(x,y)).
\]
Intuitively, the subformula $\beta(S) := \pa{x} (\ex{y} S(x,y) \wedge M(x,y))$ calculates the value \linebreak $\prod_{i=1}^n A[i, \sigma(i)]$ whenever $S$ encodes a permutation $\sigma$.
Moreover, the subformula $\op{permut}(S) \cdot \beta(S)$ returns $\beta(S)$ when $S$ is a permutation, and returns $0$ otherwise (i.e. $\op{permut}(S)$ behaves like a filter). 
Finally, the second order sum aggregates these values iterating over all binary relations and calculating the permanent of the matrix.
We would like to finish with this example by highlighting the similarity of $\alpha_3$ to the permanent formula. 
Indeed, an advantage of $\qso$-formulae is that the first- and second-order quantifiers in the quantitative level naturally reflect the operations used to define mathematical formulae. \qed
\end{exa}

We consider several fragments or extensions of $\qso$, which are obtained by restricting the syntax of the Boolean formulae or the use of the quantitative quantifiers.
In this regard, we denote by $\qfo$ the fragment of $\qso$ where second-order sum and product are not allowed. 
For instance, for the $\qso$-formulae defined in Example \ref{ex:cliques}, we have that $\alpha_1$ is in $\qfo$ and $\alpha_2$ is not.
Another interesting fragment of $\qso$ consists of the $\qso$-formulae where only sum operators and sum quantifiers are allowed. 
Formally, we denote by $\eqso$ the fragment of $\qso$ where first- and second-order products (i.e. $\pa{x}$ and $\pa{X}$) are not allowed.
For example, $\alpha_1$ and $\alpha_2$ in Example \ref{ex:cliques} are formulae of $\eqso$, while $\alpha_3$ in Example \ref{exa-perm} is not. 
We also consider fragments of $\qso$ by further restricting the Boolean part of the logic.
If $\LL$ is a fragment of $\so$, then we define the quantitative logic $\qso(\LL)$ to be the fragment of $\qso$ obtained by restricting $\varphi$ in \eqref{syntax} to be a formula in $\LL$. Moreover, we also restrict other fragments of $\qso$ by using the same idea. 
For example, we define $\qfo(\fo)$ to be the fragment of $\qfo$ obtained by restricting $\varphi$ in \eqref{syntax} to be an $\fo$-formula, and likewise for $\eqso(\fo)$.

In the following section, we use different fragments or extensions of $\qso$ to capture counting complexity classes. But before doing this, we show the connection of $\qso$ to previous frameworks for defining functions over relational structures.

@@

In this section, we discuss some previous frameworks proposed in the literature and how they differ from our approach.
We start by discussing the connection between $\qso$ and weighted logics (WL)~\cite{DrosteG07}. 
As it was previously discussed, $\qso$ is a fragment of WL.
The main difference is that we restrict the semiring used in WL to natural numbers in order to study counting complexity classes.
Another difference between WL and our approach is that, to the best of our knowledge, this is the first paper to study weighted logics over general relational signatures, in order  to do descriptive complexity for counting complexity classes. 
Previous works on WL usually restrict the signature of the logic to strings, trees, and other specific structures (see \cite{droste2009handbook} for more examples), and they did not study the logic over general structures. 
Furthermore, in this paper we propose further extensions for $\qso$ (see Section~\ref{sec:beyond}) which differ from previous approaches in WL.

>>>>

The syntax of QSO formulae is divided in two levels. 
The first level is composed by $\so$-formulae over $\R$ (called Boolean formulae) and the second level is made by counting operators of addition and multiplication. 
For this reason, the quantifiers in $\so$ (e.g. $\exists x$ or $\exists X$) are called Boolean quantifiers and the quantifiers that make use of addition and multiplication (e.g. $\Sigma x$ or $\Pi X$) are called {\em quantitative quantifiers}.
Furthermore, $\Sigma x$ and $\Sigma X$ are called first- and second-order sum, and $\Pi x$ and $\Pi X$ are called first- and second-order product, respectively.
This division between Boolean and quantitative level is essential for understanding the difference between the logic and the quantitative part. 
Furthermore, this will allow us later to parametrize both levels of the logic in order to capture different counting complexity classes.

@@

Suppose now that we want to count the number of cliques in~$\fG$. We can define this function with the following formula:
\[
\alpha_2 \ := \ \sa{X} \clique(X),
\] 
where $\clique(X) := \fa{x} \fa{y} ((X(x) \wedge X(y) \wedge x \neq y)  \rightarrow E(x,y))$.
In the Boolean sub-formula of $\alpha_2$ we check whether $X$ is a clique, and with the sum quantifier we add one for each clique in $\fG$. 
But in contrast to $\alpha_1$, 
in $\alpha_2$ we need a second-order quantifier in the quantitative level.
This is according to the
complexity of evaluating each formula:
$\alpha_1$ defines an $\fp$-function while $\alpha_2$ defines a $\shp$-complete function. \qed
\end{exa}
\begin{exa}\label{exa-perm}
For an example that includes multiplication, let $\bM = \{M(\cdot,\cdot),<\}$ be a vocabulary for storing 0-1 matrices; in particular, a structure $\fM$ over $\bM$ encodes a 0-1 matrix $A$ as follows: if $A[i,j] = 1$, then $M(i,j)$ is true, otherwise $M(i.j)$ is false.
Suppose now that we want to compute the permanent of an $n$-by-$n$ 0-1 matrix $A$, that is:
\begin{align*}
\op{perm}(A) &= \sum_{\sigma \in S_n} \prod_{i=1}^n A[i, \sigma(i)],  
\end{align*}
where $S_n$ is the set of all permutations over $\{1, \ldots, n\}$.
The permanent is a fundamental function on matrices that has found many applications;
% in different areas,
%~\cite{permanent-applications},
in fact, showing that this function is hard to compute was one of the main motivations behind the definition of the class $\shp$~\cite{Valiant79}.
%and it was one of the first function that was shown to be difficult for counting~\cite{Valiant79} (i.e. $\shp$-complete).

To define the permanent of a 0-1 matrix in $\qso$, assume that for a binary relation symbol $S$, $\op{permut}(S)$ is an $\fo$-formula that is true if, and only if, $S$ is a permutation, namely, a total bijective function (the definition of $\op{permut}(S)$ is straightforward).
%\martin{abreviar $\op{permutation}$ a $\op{permut}$ ahorra un par de lineas}
Then the following is a $\qso$-formula defining the permanent of a matrix:
\[
\alpha_3 := \sa{S} \op{permut}(S) \cdot \pa{x} (\ex{y} S(x,y) \wedge M(x,y)).
\]
Intuitively, the subformula $\beta(S) := \pa{x} (\ex{y} S(x,y) \wedge M(x,y))$ calculates the value \linebreak $\prod_{i=1}^n A[i, \sigma(i)]$ whenever $S$ encodes a permutation $\sigma$.
Moreover, the subformula $\op{permut}(S) \cdot \beta(S)$ returns $\beta(S)$ when $S$ is a permutation, and returns $0$ otherwise (i.e. $\op{permut}(S)$ behaves like a filter). 
Finally, the second order sum aggregates these values iterating over all binary relations and calculating the permanent of the matrix.
We would like to finish with this example by highlighting the similarity of $\alpha_3$ with the permanent formula. 
Indeed, an advantage of $\qso$-formulae is that the first- and second-order quantifiers in the quantitative level naturally reflect the operations used to define mathematical formulae. \qed
\end{exa}

We consider several fragments or extensions of $\qso$, which are obtained by restricting the syntax of the Boolean formulae or the use of the quantitative quantifiers.
In this direction, we denote by $\qfo$ the fragment of $\qso$ where second-order sum and product are not allowed. 
For instance, for the $\qso$-formulae defined in Example \ref{ex:cliques}, we have that $\alpha_1$ is in $\qfo$ and $\alpha_2$ is not.
Another interesting fragment of $\qso$ consists of the $\qso$-formulae where only sum operators and sum quantifiers are allowed. 
Formally, we denote by $\eqso$ the fragment of $\qso$ where first- and second-order products (i.e. $\pa{x}$ and $\pa{X}$) are not allowed.
For example, $\alpha_1$ and $\alpha_2$ in Example \ref{ex:cliques} are formulae of $\eqso$, while $\alpha_3$ in Example \ref{exa-perm} is not. 
We also consider fragments of $\qso$ by further restricting the Boolean part of the logic.
If $\LL$ is a fragment of $\so$, then we define the quantitative logic $\qso(\LL)$ to be the fragment of $\qso$ obtained by restricting $\varphi$ in \eqref{syntax} to be a formula in $\LL$. Moreover, we also restrict other fragments of $\qso$ by using the same idea. 
For example, we define $\qfo(\fo)$ to be the fragment of $\qfo$ obtained by restricting $\varphi$ in \eqref{syntax} to be an $\fo$-formula, and likewise for $\eqso(\fo)$.

In the following section, we use different fragments or extensions of $\qso$ to capture counting complexity classes. But before doing this, we show the connection of $\qso$ with previous frameworks for defining functions over relational structures.

@@

In this section, we discuss some previous frameworks proposed in the literature and how they differ from our approach.
We start by discussing the connection between $\qso$ and weighted logics (WL)~\cite{DrosteG07}. 
As it was previously discussed, $\qso$ is a fragment of WL.
The main difference is that we restrict the semiring used in WL to natural numbers in order to study counting complexity classes.
Another difference of WL with our approach is that, to the best of our knowledge, this is the first paper to study weighted logics over general relational signatures, in order  to do descriptive complexity for counting complexity classes. 
Previous works on WL usually restrict the signature of the logic to strings, trees, and other specific structures (see \cite{droste2009handbook} for more examples), and they did not study the logic over general structures. 
Furthermore, in this paper we propose further extensions for $\qso$ (see Section~\ref{sec:beyond}) which differ from previous approaches in WL.

### 2018-12-27 LMCS/sendrev/main.tex
carpeta con version send corregida

%% mandatory lists of keywords 
\keywords{MANDATORY list of keywords}

%% read in additional TeX-packages or personal macros here:
%% e.g. \usepackage{tikz}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,mathabx}
\usepackage{stmaryrd}
\usepackage[textwidth=2cm,textsize=small]{todonotes}
\usepackage{enumitem}
\input{macros}

%% define non-standard environments BEYOND the ones already supplied 
%% here, for example
%\theoremstyle{plain}\newtheorem{satz}[thm]{Satz} %\crefname{satz}{Satz}{S\"atze}
%% Do NOT replace the proclamation environments lready provided by
%% your own.

@@

\title[Counting]{Descriptive Complexity for Counting Complexity Classes}
\titlecomment{{\lsuper*}OPTIONAL comment concerning the title, \eg, 
  if a variant or an extended abstract of the paper has appeared elsewhere.}

@@

\newpage
\input{todo-list}

>>>>

%% mandatory lists of keywords 
\keywords{Function complexity clases, descriptive complexity}
\ACMCCS{
[{\bf Theory of computation}]:~Logic---Finite Model Theory
[{\bf Theory of computation}]:~Computational complexity and cryptography---Complexity classes}

%% read in additional TeX-packages or personal macros here:
%% e.g. \usepackage{tikz}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,mathabx}
\usepackage{stmaryrd}
\input{macros}
%% define non-standard environments BEYOND the ones already supplied 
%% here, for example
\theoremstyle{plain}\newtheorem{satz}[thm]{Satz} %\crefname{satz}{Satz}{S\"atze}
%% Do NOT replace the proclamation environments lready provided by
%% your own.

@@

\title[Counting]{Descriptive Complexity for Counting Complexity Classes}
%\titlecomment{{\lsuper*}OPTIONAL comment concerning the title, \eg, 
%  if a variant or an extended abstract of the paper has appeared elsewhere.}

@@

### 2018-12-27 LMCS/sendrev/proof-theo-pi1-pnf.tex
carpeta con version send corregida

% !TeX root = main.tex

@@

Second, we prove that the formula $\alpha_{1} = 2$ (i.e. $\sem{\alpha_{1}}$ is the constant function $2$) is not equivalent to any formula in $\loge{1}$-PNF. 
Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\, \varphi(\bar{X},\bar{x},\bar{y})$ in $\loge{1}$-PNF that is equivalent to $\alpha_1$. 
First, if $\length{\bar{X}} = \length{\bar{x}} = 0$, then the function defined by $\alpha$ is never greater than 1. 
Therefore, suppose that $\length{\bar{X}} > 0$ or $\length{\bar{x}} > 0$, and consider some ordered structure $\A$. 
Since $\sem{\alpha}(\A) = 2$, there exist at least two assignments $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, $(\bar{B}_2,\bar{b}_2,\bar{a}_2)$ to $(\bar{X},\bar{x},\bar{y})$ such that for $i\in\{1,2\}$: $\A\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$. 
Now consider the ordered structure $\A'$ that is obtained by taking the disjoint union of $\A$ twice. 
Indeed, each half of $\A'$ is isomorphic to $\A$. 
Note that $\A'\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$ for $i = 1,2$ and there exists a third assignment $(\bar{B}_1',\bar{b}_1',\bar{a}_1')$ that is isomorphic to $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, in the other half of the structure, such that $\A'\models\varphi(\bar{B}_1',\bar{b}_1',\bar{a}_1')$. 
As a result, we have that $\sem{\alpha}(\A') \geq 3$ which leads to a contradiction.

For the last part of the proof, we show that if $\LL$ contains $\logu{1}$ and is closed under conjunction and disjunction, then for every formula $\alpha$ in $\eqso(\LL)$ there exists an equivalent formula in $\LL$-PNF. 
Similarly as in the proof of Theorem~\ref{theo-pnf-snf}, we show a recursive function $\tau$ that produces such a formula. 
Assume that $\alpha = \sum_{i = 1}^n \alpha_i$ is in $\LL$-SNF where each $\alpha_i$ is in $\LL$-PNF. 
Without loss of generality, we assume that each $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi_i(\bar{X},\bar{x})$ with $\length{\bar{X}} > 0$ and $\length{\bar{x}} > 0$. 
If that is not the case, we can replace each $\alpha_i$ by the equivalent formula
$$
\sa{\bar{X}} \sa{Y}\sa{\bar{x}}\sa{y}(\varphi_i(\bar{X},\bar{x})\wedge \fa{z} Y(z) \wedge \fa{z} z \leq y).
$$
Now we begin describing the function $\tau$. 
If $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$, then the formula is already in $\LL$-PNF so we define $\tau(\alpha) = \alpha$. 
If $\alpha = \alpha_1 + \alpha_2$, then we assume that $\tau(\alpha_1) = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ and $\tau(\alpha_2) = \sa{\bar{Y}}\sa{\bar{y}}\psi(\bar{Y},\bar{y})$. 
Our construction works by identifying a ``first'' assignment for both $(\bar{X},\bar{x})$ and $(\bar{Y},\bar{y})$ and a ``last'' assignment for both $(\bar{X},\bar{x})$ and $(\bar{Y},\bar{y})$ using the following formulas:
\begin{align*}
\gamma_{\text{first}}(\bar{X},\bar{x}) & \; = \;  \bigwedge_{i = 1}^{\length{\bar{X}}} \fa{\bar{z}}\neg X_i(\bar{z}) \wedge \fa{\bar{z}}(\bar{x}\leq\bar{z}), \\
\gamma_{\text{last}}(\bar{X},\bar{x}) & \; = \;  \bigwedge_{i = 1}^{\length{\bar{X}}} \fa{\bar{z}} X_i(\bar{z}) \wedge \fa{\bar{z}}(\bar{z}\leq\bar{x}).
\end{align*}
Similarly, we can define the formulas $\gamma_{\text{first}}(\bar{Y},\bar{y})$ and $\gamma_{\text{last}}(\bar{Y},\bar{y})$.
In other words, the ``first'' assignment is the one where every second-order predicate is empty and the first-order assignment is the lexicographically smallest, and the ``last'' assignment is the one where every second-order predicate is full and the first-order assignment is the lexicographically greatest. 
We also need to identify the assignments that are not first and the ones that are not last. 
We do this by negating the two formulas above and grouping together the first-order variables:
\begin{align*}
\gamma_{\text{not-first}}(\bar{X},\bar{x}) & \; = \; \ex{\bar{z}}(\bar{z}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{z}_i)), \\
\gamma_{\text{not-last}}(\bar{X},\bar{x}) & \; = \; \ex{\bar{z}}(\bar{x} < \bar{z}_0 \vee \bigvee_{i = 1}^{\length{\bar{X}}}\neg X(\bar{z}_i)),
\end{align*}
where $\bar{z} = (\bar{z}_0,\bar{z}_1,\ldots,\bar{z}_{\length{\bar{X}}})$. Then the following formula is equivalent to $\alpha$:
\begin{align}
\sa{\bar{X}}\sa{\bar{x}}\sa{\bar{Y}}\sa{\bar{y}}[&(\varphi(\bar{X},\bar{x})\wedge\gamma_{\text{not-first}}(\bar{X},\bar{x})\wedge\gamma_{\text{first}}(\bar{Y},\bar{y}))\vee \label{eq:partition1} \\
&(\varphi(\bar{X},\bar{x})\wedge\gamma_{\text{first}}(\bar{X},\bar{x})\wedge\gamma_{\text{last}}(\bar{Y},\bar{y}))\vee \label{eq:partition2}\\
&(\psi(\bar{Y},\bar{y})\wedge\gamma_{\text{first}}(\bar{X},\bar{x})\wedge\gamma_{\text{not-last}}(\bar{Y},\bar{y}))\vee \label{eq:partition3}\\
&(\psi(\bar{Y},\bar{y})\wedge\gamma_{\text{last}}(\bar{X},\bar{x})\wedge\gamma_{\text{last}}(\bar{Y},\bar{y}))]. \label{eq:partition4}
\end{align}

>>>>

%!TEX root = main.tex

@@

Second, we prove that the formula $\alpha_{1} = 2$ (i.e. $\sem{\alpha_{1}}$ is the constant function $2$) is not equivalent to any formula in $\loge{1}$-PNF. 
Suppose that there exists some formula $\alpha = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\, \varphi(\bar{X},\bar{x},\bar{y})$ in $\loge{1}$-PNF that is equivalent to $\alpha_1$. 
First, if $\length{\bar{X}} = \length{\bar{x}} = 0$, then the function defined by $\alpha$ is never greater than 1. 
Therefore, suppose that $\length{\bar{X}} > 0$ or $\length{\bar{x}} > 0$, and consider some ordered structure $\A$. 
Since $\sem{\alpha}(\A) = 2$, there exist at least two assignments $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, $(\bar{B}_2,\bar{b}_2,\bar{a}_2)$ to $(\bar{X},\bar{x},\bar{y})$ such that for both, $\A\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$. 
Now consider the ordered structure $\A'$ that is obtained by taking the disjoint union of $\A$ twice. 
Indeed, each half of $\A'$ is isomorphic to $\A$. 
Note that $\A'\models\varphi(\bar{B}_i,\bar{b}_i,\bar{a}_i)$ for $i = 1,2$ and there exists a third assignment $(\bar{B}_1',\bar{b}_1',\bar{a}_1')$ that is isomorphic to $(\bar{B}_1,\bar{b}_1,\bar{a}_1)$, in the other half of the structure, such that $\A'\models\varphi(\bar{B}_1',\bar{b}_1',\bar{a}_1')$. 
As a result, we have that $\sem{\alpha}(\A') \geq 3$ which leads to a contradiction.

For the last part of the proof, we show that if $\LL$ contains $\logu{1}$ and is closed under conjunction and disjunction, then for every formula $\alpha$ in $\eqso(\LL)$ there exists an equivalent formula in $\LL$-PNF. 
Similarly to Theorem~\ref{theo-pnf-snf}, we show a recursive function $\tau$ that produces such a formula. 
Assume that $\alpha = \sum_{i = 1}^n \alpha_i$ is in $\LL$-SNF where each $\alpha_i$ is in $\LL$-PNF. 
Without loss of generality, we assume that each $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi_i(\bar{X},\bar{x})$ with $\length{\bar{X}} > 0$ and $\length{\bar{x}} > 0$. 
If that is not the case, we can replace each $\alpha_i$ by the equivalent formula
$$
\sa{\bar{X}} \sa{Y}\sa{\bar{x}}\sa{y}(\varphi_i(\bar{X},\bar{x})\wedge \fa{z} Y(z) \wedge \fa{z} z \leq y).
$$
Now we begin describing the function $\tau$. 
If $\alpha = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$, then the formula is already in $\LL$-PNF so we define $\tau(\alpha) = \alpha$. 
If $\alpha = \alpha_1 + \alpha_2$, then we assume that $\tau(\alpha_1) = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ and $\tau(\alpha_2) = \sa{\bar{Y}}\sa{\bar{y}}\psi(\bar{Y},\bar{y})$. 
Our construction works by identifying a ``first'' assignment for both $(\bar{X},\bar{x})$ and $(\bar{Y},\bar{y})$ and a ``last'' assignment for both $(\bar{X},\bar{x})$ and $(\bar{Y},\bar{y})$ using the following formulas:
\begin{align*}
\gamma_{\text{first}}(\bar{X},\bar{x}) & \; = \;  \bigwedge_{i = 1}^{\length{\bar{X}}} \fa{\bar{z}}\neg X_i(\bar{z}) \wedge \fa{\bar{z}}(\bar{x}\leq\bar{z}), \\
\gamma_{\text{last}}(\bar{X},\bar{x}) & \; = \;  \bigwedge_{i = 1}^{\length{\bar{X}}} \fa{\bar{z}} X_i(\bar{z}) \wedge \fa{\bar{z}}(\bar{z}\leq\bar{x}).
\end{align*}
Similarly, we can define the formulas $\gamma_{\text{first}}(\bar{Y},\bar{y})$ and $\gamma_{\text{last}}(\bar{Y},\bar{y})$.
In other words, the ``first'' assignment is the one where every second-order predicate is empty and the first-order assignment is the lexicographically smallest, and the ``last'' assignment is the one where every second-order predicate is full and the first-order assignment is the lexicographically greatest. 
We also need to identify the assignments that are not first and the ones that are not last. 
We do this by negating the two formulas above and grouping together the first-order variables:
\begin{align*}
\gamma_{\text{not-first}}(\bar{X},\bar{x}) & \; = \; \ex{\bar{z}}(\bar{z}_0 < \bar{x} \vee \bigvee_{i = 1}^{\length{\bar{X}}}X(\bar{z}_i)), \\
\gamma_{\text{not-last}}(\bar{X},\bar{x}) & \; = \; \ex{\bar{z}}(\bar{x} < \bar{z}_0 \vee \bigvee_{i = 1}^{\length{\bar{X}}}\neg X(\bar{z}_i)),
\end{align*}
where $\bar{z} = (\bar{z}_0,\bar{z}_1,\ldots,\bar{z}_{\length{\bar{X}}})$. Then the following formula is equivalent to $\alpha$:
\begin{align}
\sa{\bar{X}}\sa{\bar{x}}\sa{\bar{Y}}\sa{\bar{y}}[&(\varphi(\bar{X},\bar{x})\wedge\gamma_{\text{not-first}}(\bar{X},\bar{x})\wedge\gamma_{\text{first}}(\bar{Y},\bar{y}))\vee \label{eq:partition1} \\
&(\varphi(\bar{X},\bar{x})\wedge\gamma_{\text{first}}(\bar{X},\bar{x})\wedge\gamma_{\text{last}}(\bar{Y},\bar{y}))\vee \label{eq:partition2}\\
&(\psi(\bar{Y},\bar{y})\wedge\gamma_{\text{first}}(\bar{X},\bar{x})\wedge\gamma_{\text{not-last}}(\bar{Y},\bar{y}))\vee \label{eq:partition3}\\
&(\psi(\bar{Y},\bar{y})\wedge\gamma_{\text{last}}(\bar{X},\bar{x})\wedge\gamma_{\text{last}}(\bar{Y},\bar{y}))]. \label{eq:partition4}
\end{align}

### 2018-12-27 LMCS/sendrev/proof-theo-pnf-snf.tex
carpeta con version send corregida

Recall that a formula in $\eqso(\LL)$ is defined by the following grammar:
\[
\alpha = \varphi \ \mid \ s \ \mid \ (\alpha + \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \sa{X} \alpha
\]
where $\varphi$ is a formula in $\LL$ and $s\in\nat$. 
To find an equivalent formula in $\LL$-SNF for every $\alpha \in \eqso(\LL)$, we give a recursive function $\tau$ such that $\tau(\alpha)$ is in $\LL$-SNF and $\tau(\alpha) \equiv \alpha$. 
Specifically, if $\alpha = \varphi$, define $\tau(\alpha) = \alpha$ and if $\alpha = s$, define $\tau(\alpha) = (\top \add \overset{\text{$s$-times}}{\ldots} \add \top)$.  If $\alpha = (\alpha_1 + \alpha_2)$, define $\tau(\alpha) = (\tau(\alpha_1) + \tau(\alpha_2))$ and if $\alpha = \sa{x}\beta$, define $\tau(\alpha) = \sum_{i = 1}^{k}\sa{x}\beta_i$ where $\tau(\beta) = \sum_{i = 1}^{k}\beta_i$ for some $k$ and each $\beta_i$ is in $\LL$-PNF. Finally, if $\alpha = \sa{X}\beta$, then we proceed analogously as in the previous case. This covers all possible cases for $\alpha$ and we conclude the proof by taking $\tau(\alpha)$ as the desired rewrite of $\alpha$.

>>>>

Recall that a formula in $\eqso(\LL)$ is defined by the following grammar:
\[
\alpha = \varphi \ \mid \ s \ \mid \ (\alpha + \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \sa{X} \alpha
\]
where $\varphi$ is a formula in $\LL$ and $s\in\nat$. 
To find an equivalent formula in $\LL$-SNF for every $\alpha \in \eqso(\LL)$, we give a recursive function $\tau$ such that $\tau(\alpha)$ is in $\LL$-SNF and $\tau(\alpha) \equiv \alpha$. 
Specifically, if $\alpha = \varphi$, define $\tau(\alpha) = \alpha$; 
if $\alpha = s$, define $\tau(\alpha) = (\top \add \overset{\text{$s $ times}}{\ldots} \add \top)$;
if $\alpha = (\alpha_1 + \alpha_2)$, define $\tau(\alpha) = (\tau(\alpha_1) + \tau(\alpha_2))$;
if $\alpha = \sa{x}\beta$, assume $\tau(\beta) = \sum_{i = 1}^{k}\beta_i$ such that each $\beta_i$ is in $\LL$-PNF, and define $\tau(\alpha) = \sum_{i = 1}^{k}\sa{x}\beta_i$;
and if $\alpha = \sa{X}\beta$, then we proceed analogously as in the previous case.
This covers all possible cases for $\alpha$ and we conclude the proof by taking $\tau(\alpha)$ as the desired rewrite of $\alpha$.
